---
title: "Analysis"
format: html
---

## Goals for this notebook
1. Investigate citation numbers over time
    i) What is the year with the most citations?
    ii) What is the month with the most citations?
    iii) What is the average number of citations for each month since 2016?
2. Investigate most common citation types
    i) What are the most common citation types of all time?
    ii) What are the most common citation types by year?
3. Judgement dispositions
    i) What is the most common judgement disposition?
   ii) How many guilty cases per year?
    iii) What is the month with the most guilty dispositions percentage?
    iv) What is the average percentage of guilty dispositions for each month since 2016?
4. People Analysis
    i) Who has received multiple citations/fines? 
        A. Who are the most heavily fined people?
    ii) Who has served jail time?
    iii) How many people still owe fines to the city?
    iv) How many people have paid their fines?
    v) How many people have had their fines dismissed? 
        A. How many of these people have been fined multiple times?

## Set up
```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
```

## Importing

```{r}
unhoused <- read_rds("data-processed/unhoused-clean.rds") # ask dominic if there's outside houston data in this set

unhoused %>% glimpse()
```

## 1. Citation numbers over time

### 1.1. Year with most citations

What is the year with the most citations?

```{r}
unhoused |>
  filter(off_yr < 2025) |> #
  group_by(year(offense_date)) |> 
  summarize(yearly_offenses = n()) |> 
  arrange() 
```

#### 1.1.1 Data takeaways

The year with the highest number of citations is 2016 with 5796, followed by 2020 with 5019. The year with the lowest number of citations is 2018, at 1385.	

### 1.2 Highest citation month

What's the month with the most citations from 2016 to 2024?

```{r}
unhoused |> 
  #filter(offense_date >= "2023-06-01") |> # i'm looking 6 months before Whitmire's term started to see if there is an uptick in citations after Jan. 2024.
  group_by(off_yr,
           off_mo_label) |> 
  summarise(
    yr_mo_off = sum(n())) |> 
  arrange(yr_mo_off)
```

#### 1.2.1 Data takeaways

The month with the highest citation number since 2016 is Jan. 2021, with 797 citations. The month with the highest citation number since June 2023 is Feb. 2024 with 294 citations, closely followed by Jun. 2024 with 280 citations.

The month with the lowest citation number since 2016 is The month with the lowest citation number since June 2023 is July 2023, with 66 citations. This is followed by June 2023 with 85.

### 1.3 Avg monthly citation

What is the average monthly citation for each month from 2016 to 2024?

```{r}
unhoused |> 
  filter(off_yr < 2025) |> 
  group_by(off_yr,
           off_mo_label) |> 
  summarise(
    yr_mo_off = sum(n())
  ) |> 
  group_by(off_mo_label) |> 
  summarise(avg_mo_off = mean(yr_mo_off) %>% 
            round(1))
```

#### 1.3.1 Data takeaways

Considering every month from 2016 to 2024, the month with the highest average monthly citations is January. The month with the lowest is November.

## 2. Citation types

### 2.1 Most common

Finding the most common violation types.

```{r}
unhoused |> 
  group_by(violation_code, violation_description) |> 
  summarise(violations = n()) |> 
  arrange(desc(violations))
```

#### 2.1.1 Data takeaways

There are seven types of violations that have been given since 2016. Here is what they mean and their frequency:

1. **CC921 has 470 citations**
    i) Sidewalk obstruction: OBSTRUCT SIDEWALK BY (PLACING/DEPOSITING/PERMIT PERSON UNDER CONTROL) OBJECT (BOX/MAT./VEH./ETC.)
2. **CC940 has 14865 citations**
    i) Sidewalk obstruction: SIT/LIE DOWN ON (BLANKET/STOOL) PLACED ON SIDEWALK B/W 7:00 AM AND 11:00 PM IN A PROHIBITED AREA
3. **CC941 has 13478 citations**
    i) Sidewalk obstruction: PLACE (BED MAT./PERSONAL POSS.) ON SIDEWALK B/W THE HRS OF 7:00 A.M. & 11:00P.M. IN A PROHIBITED AREA
4. **CC942 has 32 citations**
    i) 	Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK WITHOUT A PERMIT
5. **CC943 has 2 citations**
    i) Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK BEYOND SCOPE OF PERMIT
6. **CC948 has 72 citations**
    i) Sidewalk obstruction: PLACE OR ALLOW OBSTRUCTION ON SIDEWALK
7. **CC753 has 5 citations**
    i) Selling goods in public area: STREET VENDOR (EXPOSE FOR SALE) (SELL) GOODS, TO-WIT:...ON A PUBLIC (SIDEWALK) (STREET) (PROPERTY)
  
The most common citation since 2016 is CC940 with 14865 offenses, with CC941 as a runner-up at 13478 offenses. The next most common is CC291 with only 470 offenses.

### 2.2 Most common by year

```{r}
unhoused |> 
  group_by(off_yr,
           violation_code)|> 
  summarise(violations = n()) |> 
  arrange(desc(off_yr))
```

#### 2.2.1 Data takeaways

All five of the CC753 citations, all 32 of CC942 citations, both of the two CC943 citations and all 72 of CC948 citations have been given in 2025.

## 3. Judgment dispositions




















